# Continuous Distributions {#continuousdist}

## Uniform Distribution

### Definition

::: {.callout-note appearance="simple"}

## Definition

The random variable $X$ has a **uniform distribution** on the interval $(a, b)$ if its density function is given by:
$$
f(x) = \begin{cases}
\frac{1}{b-a}, & a \leq x \leq b \\
0,				& \text{otherwise.}
\end{cases}
$$

The distribution function of $X$ is: 
\begin{align*}
F(x) &=  
\begin{cases}
0 & \text{for } x < a  \\
\frac{x-a}{b-a}, & a \leq x \leq b \\
1 & \text{for } x > b
\end{cases}
\end{align*}

\textbf{Mean: }$E[X] = \frac{(a+b)}{2}$

\textbf{Variance:} $Var[X] = \frac{(b-a)^2}{12}$ 

::: 

```{r uniform_pdf_cdf, echo=FALSE, fig.height=5, fig.width=10, out.width='100%'}
par(mfrow=c(1,2))
x <- seq(0, 10, 0.001)
plot(x, dunif(x, 5, 8), type="l", main="Density Function", ylab="f(x)")
plot(x, punif(x, 5, 8), type="l", main="Distribution Function", ylab="F(x)")
```

### Uniform Distribution in R

`dunif(x, min, max)`

`punif(x, min, max)`

`qunif(Fx, min, max)`

`runif(numSims, min, max)`

### Example 1: Uniform distribution {.unnumbered}

Let $X \sim U(0, 2)$. Find $P(X  \leq 0.5)$.

```{r, echo=TRUE}
(0.5-0)/(2-0)
punif(0.5, 0, 2)
```

### Example 2: Uniform distribution {.unnumbered}

A farmer living in western Nebraska has an irrigation system to provide water for crops, primarily corn, on a large farm. Although he has thought about buying a backup pump, he has not done so. If the pump fails, delivery time $X$ for a new pump to arrive is uniformly distributed over the interval from 1 to 4 days. The pump fails. It is a critical time in the growing season in that the yield will be greatly reduced if the crop is not watered within the next 3 days. 

(Scheaffer and Young, 2010, ex 5.7, page 213)

a) Assuming that the pump is ordered immediately and that installation time is negligible, what is the probability that the farmer will suffer major yield loss?

*Solution*
$P(X > 3)  = \int_3^4 \frac{1}{3} dx = \frac{1}{3}$

b) Sketch the density function and distribution function of $X$

```{r, farmer_crops_pdfcdf, out.width='50%', echo=FALSE, fig.align='center'}
x <- seq(0, 5, 0.01)
plot(x, dunif(x, 1, 4), ylab="f(x)", pch="-", main= "X ~ Uniform(1,4)")
plot(x, punif(x, 1, 4), ylab="F(x)", type="l", main= "X ~ Uniform(1,4)")
```

### Example 3: Uniform distribution {.unnumbered}

Derive the distribution function of the uniform random variable.

*Solution*

The cdf can be derived as follows:
\begin{align*}
F(x) &= \int_a^x f(t) dt  = \int_a^x \frac{1}{b-a} dt \\
&= \left[\frac{t}{b-a} \right]_a^x \\
%&= \frac{x}{d-c} - \frac{c}{d-c}  \\
&=  \frac{x-a}{b-a}, \quad  a \leq x \leq b
\end{align*}
	
### Properties of the Uniform Distribution

For a subinterval $(c,d)$ of $(a,b)$, then:
\begin{align*}
P(c \leq X \leq d) & = \int_{c}^{d} \frac{1}{b-a} dx  \\
& = \left. \frac{x}{b-a} \right|^d_c \\
& =  \frac{d-c}{b-a}\\
\end{align*}

Special case: **uniform $[0,1]$ random variable**.

$$
f(x) = \begin{cases}
1 & 0 < x < 1 \\
0 & \text{otherwise}
\end{cases}
$$

## Exponential Distribution

### Definition

::: {.callout-note appearance="simple"}

## Definition

A random variable $X$ has an **exponential distribution**, $X \sim \textrm{Exp}(\lambda)$, if and only if it has density function:

\begin{equation*}
f(x) = 
\begin{cases}
\lambda e^{-\lambda x} & \qquad \text{where } x \geq 0, \lambda > 0 \\
0 & \qquad \text{otherwise}
\end{cases}
\end{equation*}

The distribution function of $X$ is:
\begin{equation*}
F(x) = 
\begin{cases}
0, & \qquad x < 0 \\
\int_{0}^{x} \lambda e^{-\lambda u} du = 1 - e^{-\lambda x}, & \qquad x \geq 0
\end{cases}
\end{equation*}

\textbf{Mean: }$E[X] =1/\lambda$

\textbf{Standard Deviation: } $Std(X) =1/\lambda$

\textbf{Variance:} $Var[X] = 1/\lambda^2$ \\

:::

### Example 4: Applications of Exponential Random Variables {.unnumbered}

- Waiting time between the arrival of successive customers
- Time to failure of an electronic component. 
- Time between successive emissions of particles from a radioactive substances

```{r exponential_pdf_cdf, echo=FALSE, fig.height=5, fig.width=10, out.width='100%'}

par(mfrow=c(1,2))

x_seq <- seq(0, 10, 0.001)

plot(x_seq, dexp(x_seq, 2), type="l", main="Density Function", ylab="f(x)")
plot(x_seq, pexp(x_seq, 2), type="l", main="Distribution Function", ylab="F(x)")

```

### Exponential Distribution in R

`dexp(x, lambda)`

`pexp(x, lambda)`

`qexp(Fx, lambda)`

`rexp(numSims, lambda)`

### Example 5: Exponential Distribution {.unnumbered}

Consider an exponential random variable $X$ with mean 2. Find $P(X > 3)$.

*Solution*

```{r, echo=TRUE}
lambda <- 1/2
1 - (1 - exp(-lambda*3))
1 - pexp(3, lambda)
```

### Example 6: Exponential Distribution {.unnumbered}

A sugar refinery has three processing plants, all of which receive raw sugar in bulk.
The amount of sugar that one plant can process in one day can be modeled as having
an exponential distribution with a mean of 4 tons for each of the three plants. 
(Scheaffer and Young, 2010, ex5.9, page 220)

a) What is the probability that a plant will process more than 4 tons on a given day?

*Solution*

$$X \sim \text{Exponential(0.25)}$$

\begin{align*}
P(X > 4) & = 1 - (1- e^{-0.25(4)})\\
& = 0.37
\end{align*}

```{r, echo=TRUE}
1-pexp(4, 0.25)
```

b) If the plants operate independently, find the probability that exactly two of the three plants will process more than 4 tons on a given day. 

*Solution*

$$N \sim \text{Binomial}(3, p)$$

where $p = P(X > 4) = 0.37$

\begin{align*}
P(N = 2) & = 
\binom{3}{2} (0.37)^2 (0.63)^1 \\
& = 0.2587
\end{align*}

```{r, echo=TRUE}
dbinom(2, 3, 0.37)
```


c) If the plants operate independently, find the probability that an odd number of plants
will process more than 4 tons on a given day. 

*Solution*

$$N \sim \text{Binomial}(3, p)$$

where $p = P(X > 4) = 0.37$

\begin{align*}
P(N = 1 \cup N=3) & = 
P(N = 1) + P(N=3) + P(N = 1 \cap N=3) \\
 & = \binom{3}{1} (0.37)^1 (0.63)^2 + \binom{3}{3} (0.37)^3 (0.63)^0 - 0
\\
& = dbinom(1, 3, 0.37)  + dbinom(3, 3, 0.37) \\
\end{align*}

```{r, echo=TRUE}
dbinom(1, 3, 0.37) + dbinom(3, 3, 0.37)
```

d) Which of the following pdfs corresponds to the amount of sugar processed by one plant?

*Solution*

The left one - $\lambda =  0.25$.

We can also check by looking at the area under the curve to the right of $4$ and comparing it to the answers above (0.37).

```{r exponential_pdf_example, echo=FALSE, fig.height=5, fig.width=10, out.width='100%'}

par(mfrow=c(1,2))

x <- seq(0, 20, 0.001)

plot(x, dexp(x, 0.25), type="l", main="Density Function") 
plot(x, dexp(x, 4), type="l", main="Density Function", xlim=c(0, 4)) 

```

### Properties

**Alternative Parameterization**

The exponential distribution is sometimes described using the mean rather than rate.

::: {.callout-note appearance="simple"}

## Properties

A random variable $X$ has an **exponential distribution** with mean $\theta$
if it has density function:

\begin{equation*}
f(x) = 
\begin{cases}
\frac{1}{\theta} e^{-  x/\theta} & \qquad \text{where } x \geq 0, \theta > 0 \\
0 & \qquad \text{otherwise}
\end{cases}
\end{equation*}

The distribution function of $X$ is:

\begin{equation*}
F(x) = 
\begin{cases}
0, & \qquad x < 0 \\
 1 - e^{- x/\theta}, & \qquad x \geq 0
\end{cases}
\end{equation*}

\textbf{Mean: }$E[X] =\theta$

\textbf{Standard Deviation:} $Std(X) =\theta$ 

\textbf{Variance:} $Var[X] = \theta^2$ 

:::

**Memoryless Property**

::: {.callout-note appearance="simple"}

## Property

If a random variable $X$ has an exponential distribution, then for all $t >0$ and $s > 0$
$$P(X > s + t | X > s ) = P(X > t)$$

This property is called the **memoryless property**.

:::

### Example 7: Memoryless Property {.unnumbered}

Suppose a customer has been waiting in line to be served for $x$ time units and would like to know the probability that he or she will be required to wait a further $t$ units of time. The calculation of this probability does not depend on the length of time already spent waiting, i.e. the distribution does not ``remember'' how long the customer has been waiting. 

### Example 8: Memoryless Property {.unnumbered}

Show that an exponential random variable with cdf $F(x) = 1 - e^{-\lambda x}, x \geq 0$ has the memoryless property.

*Solution*

\begin{align*}
P(X > s + t | X > s) 
&= P((X>s+t) \cap (X>s))/P(X>s)\\
&= P(X>s+t)/P(X>s)\\ % \qquad \text{\footnotesize If X $>$ s+t then X$>$s }\\
& =e^{-\lambda (s+t)}/e^{-\lambda s} \\ 
&= e^{-\lambda t} = P(X>t)
\end{align*}

## Gamma Distribution

### Definition

::: {.callout-note appearance="simple"}

## Definition

A random variable $X$ has a **Gamma distribution**, $X \sim \textrm{Gamma}(\alpha, \lambda)$, with shape parameter $\alpha >0$ and rate parameter $\lambda >0$ if it has a density function:

\begin{equation*}
f(x) = 
\begin{cases}
\frac{x^{\alpha-1}\lambda^\alpha  e^{-\lambda x}}{\Gamma(\alpha)}  & 
\qquad x \geq 0\\
0 & x < 0
\end{cases}
\end{equation*}

\textbf{Mean: } $E[X] = \alpha/\lambda$ 

\textbf{Standard deviation:} $Std(X) = \sqrt{\alpha}/\lambda$

\textbf{Variance: }$Var[X] = \alpha/\lambda^2$

:::

```{r gamma_pdf_cdf, echo=FALSE, fig.height=5, fig.width=10, fig.align='center', out.width='100%'}

par(mfrow=c(1,2))

x <- seq(0, 20, 0.001)
alpha <- c(0.5, 1, 2, 4)
lambda <- 1

#PDF
plot(x, dgamma(x, alpha[1], lambda), 
		ylab = "f(x)", type="l", col=1, ylim=c(0, 1.2),
         main = expression(paste("PDF: X ~ Gamma(", alpha, ",", lambda, ")")))

for(i in 2:length(alpha)){
    lines(x, dgamma(x, alpha[i], lambda), type="l", col=i)
  }


leg = paste("Gamma(", alpha, ", ", lambda, ")", sep="")
legend("topright", legend=leg, col=1:length(alpha), lty=1)

#CDF

plot(x, pgamma(x, alpha[1], lambda), 
		ylab = "F(x)", type="l", col=1,
         main = expression(paste("CDF: X ~ Gamma(", alpha, ",", lambda, ")")))

for(i in 2:length(alpha)){
    lines(x, pgamma(x, alpha[i], lambda), type="l", col=i)
  }


leg = paste("Gamma(", alpha, ", ", lambda, ")", sep="")
legend("bottomright", legend=leg, col=1:length(alpha), lty=1)

```

### Example 9: Applications of the Gamma Distribution {.unnumbered}

- Component lifetimes (a few fail early, many have an ``average'' lifetime, and a few last a very long time)
- Rainfall over a given period of time
- Survival times
- Fish lengths
- Time until failure of the $\alpha$th independent exponentially distributed component

### Gamma Distribution in R

`dgamma(x, alpha, lambda)`

`pgamma(x, alpha, lambda)`

`qgamma(Fx, alpha, lambda)`

`rgamma(numSims, alpha, lambda)`

### Example 10: Gamma Distribution in R {.unnumbered}

Consider a gamma random variable $X$ with shape parameter 2 and rate parameter 3. Find $P(X < 2.5)$.

*Solution*

```{r}
pgamma(2.5, 2, 3)
```

### Properties of the Gamma Distribution

### Distribution Function {.unnumbered}

The **cumulative distribution function** is not available in closed form, but can be evaluated numerically:

\begin{equation*}
F(x; \alpha, \lambda) = \int_{0}^{x} \frac{t^{\alpha-1}\lambda^\alpha  e^{-\lambda t}}{\Gamma(\alpha)} dt
\end{equation*}

In `R`, the distribution function can be calculated using the command:

`pgamma(x, shape = alpha, rate = lambda)`  (default)

or

`pgamma(x, shape = alpha, scale = 1/lambda)`

### Example 11: Gamma Distribution {.unnumbered}

The weekly downtime $X$ (in hours) for a certain industrial machine has approximately a gamma distribution with $\alpha$ = 3.5 and $\lambda$ = 2/3. (Adapted from Scheaffer and Young, ex5.67, pp232)

a) Find the expected value and variance of $X$

*Solution*

$E[X] = \alpha/\lambda = \frac{3.5}{2/3}$ hours

$Var[X] = \alpha/\lambda^2 = \frac{3.5}{(2/3)^2}$ hours

b) What is the probability that the weekly downtime will exceed 5 hours? Use the relevant `R` code below to compute your answer.

*Solution*

```{r downtime_gammacdf}
pgamma(seq(0, 6, 0.5), 3.5, 2/3)
pgamma(seq(0, 6, 0.5), 3.5, scale=1.5)
dgamma(seq(0, 6, 0.5), 3.5, 2/3)
```

$P(X > 5) = 1 - P(X \leq 5)$

```{r}
1 - pgamma(5, 3.5, 2/3)
```

c) Suppose the loss $L$ (in dollars) to the industrial operation as a result of this downtime is given by
$$L = 30X + 2X^2$$

Find the expected value of $L$.

*Solution*

Recall that:
$Var[X] = E[X^2] - (E[X])^2$.

For a gamma random variable:

$$E[X^2] = \alpha/\lambda^2 + \alpha^2/\lambda^2 = \alpha(1 + \alpha)/\lambda^2$$

\begin{align*}
E[L] & = E[30X + 2X^2] \\
& = 30E[X] + 2E[X^2]\\
&= 30 \left(\frac{3.5}{2/3}\right) + 2\left( \frac{3.5(1 + 3.5)}{(2/3)^2}\right)
= \$228.38 
\end{align*}

```{r}
alpha <- 3.5; lambda <- 2/3

30*alpha/lambda + 2*alpha*(1+alpha)/(2/3)^2
```

#### Relationship to Other Distributions {.unnumbered}

The Gamma distribution is closely related to a number of other distributions. 

- If $\alpha = 1$, then this gives the \textbf{exponential distribution} with parameter $\lambda$.
- If $\alpha$ is a positive integer, then this gives the \textbf{Erlang distribution}.
- If $\alpha = n/2$ where $n$ is a positive integer and $\lambda = 1/2$, then this gives a \textbf{chi-square distribution}.

#### Sum of Gamma Random Variables {.unnumbered}

::: {.callout-note appearance="simple"}

## Property

The \textbf{sum of $n$ independent gamma random variables} also has a gamma distribution.
If $X_i$ are independent gamma random variables with parameters $\alpha_i$ and $\lambda$
and 
$Y = \sum_{i=1}^n X_i$, then 

$$Y \sim \text{Gamma}(\sum_{i=1}^n\alpha_i, \lambda)$$

$$\displaystyle E[Y] =  \frac{\sum_{i=1}^n\alpha_i}{\lambda} \qquad \qquad 
\displaystyle Var[Y] = \frac{\sum_{i=1}^n\alpha_i}{\lambda^2} $$

:::

## Normal Distribution

### Definition

::: {.callout-note appearance="simple"}

## Definition

A random variable $X$ is **Normally distributed**, $X \sim N(\mu, \sigma^2)$,
 if and only if
it has probability density function:

\begin{equation*}
f_X(x) = \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}
\right)^2}, \qquad x \in \mathbb{R}, \sigma > 0
\end{equation*}

\textbf{Mean:}  $\mu$ 

\textbf{Variance:} $\sigma^2$

**Special case:** The **standard normal distribution**, denoted $Z$, has a mean of $\mu=0$ and a standard deviation of $\sigma=1$, 
$$Z \sim  N(0, 1)$$

:::


```{r normal_pdf_cdf, echo=FALSE, fig.height=5, fig.width=10, fig.align='center', out.width='100%'}

par(mfrow=c(1,2))

x <- seq(-15, 25, 0.1)

mu <- c(0, 0, 4, 10)
sigma <- c(1, 3, 5, 5)
n <- length(mu)

#PDF
plot(x, dnorm(x, mu[1], sigma[1]), 
		ylab = "f(x)", type="l", col=1, 
		xlim = c(-15, 25),
		#ylim=c(0, 1.2),
         main = expression(paste("PDF: X ~ Normal(", mu, ",", sigma^2, ")")))

for(i in 2:n){
    lines(x, dnorm(x, mu[i], sigma[i]), type="l", col=i)
  }


leg = paste("Normal(", mu, ", ", sigma^2, ")", sep="")
legend("topright", legend=leg, col=1:n, lty=1)

#CDF

plot(x, pnorm(x, mu[1], sigma[1]), 
		ylab = "F(x)", type="l", col=1,
		xlim = c(-15, 25),
         main = expression(paste("CDF: X ~ Normal(", mu, ",", sigma^2, ")")))

for(i in 2:n){
    lines(x, pnorm(x, mu[i], sigma[i]), type="l", col=i)
  }


leg = paste("Normal(", mu, ", ", sigma^2, ")", sep="")
legend("bottomright", legend=leg, col=1:n, lty=1)

```

### Example 12: Applications of Normal Distribution {.unnumbered}

- Experimental errors in scientific measurements
- Test scores on achievement tests
- Heights of individuals selected at random from a population
- Approximation to other distributions

### Normal Distribution in R

`dnorm(x, mean, sd)`

`pnorm(x, mean, sd)`

`qnorm(Fx, mean, sd)`

`rnorm(numSims, mean, sd)`

### Example 13: Normal Distribution in R {.unnumbered}

Consider a normal random variable $X$ with mean 3 and variance 0.2. Find $P(X < 2.8)$.

*Solution*

```{r}
pnorm(2.8, 3, sqrt(0.2))
```

### Relationship between Normal and Standard Normal

Suppose $X \sim N(\mu, \sigma^2)$ and $Z \sim N(0,1)$.

The random variable $X$ can be \textbf{transformed} to a standard Normal random variable as follows:

$$
Z = \frac{X - \mu}{\sigma}
$$

Then
$$
P(X \leq x) = P( Z \leq \frac{x - \mu}{\sigma})  = P( Z \leq z)
$$

In the **old days**, before `R`, statisticians had to use a book of tables to compute the probabilities. Rather than carrying around a table for every single combination of $\mu$ and $\sigma$, they would use tables for the standard Normal distribution and convert their $X$ to a $Z$ as shown above.


$$P(\mu-1\sigma \leq X \leq \mu+1\sigma) \approx 0.68 \quad \mbox{1 standard deviation from mean}$$

$$P(\mu-2\sigma \leq X \leq \mu+2\sigma) \approx 0.95 \quad \mbox{2 standard deviations from mean}$$

$$P(\mu-3\sigma \leq X \leq \mu+3\sigma) \approx 0.997 \quad \mbox{3 standard deviations from mean}$$

**R code to create plot**

```{r normal_probability, echo=TRUE, out.width='70%', fig.width = 8, fig.height =5}
mu <- 0
sigma <- 1
x <- seq(-4, 4, 0.001)

cols <- c(gray.colors(4, start = 0.1, end = 0.9, gamma = 2.2, alpha = NULL),
          gray.colors(4, start = 0.9, end = 0.1, gamma = 2.2, alpha = NULL))
colText <- c("black")


#Create plot
plot(x, dnorm(x, mu, sigma), type="l", lwd=3, ylab="f(x)", yaxt="n", xaxt="n")

#Add axis labels
axis(side=1, at=-3:3, labels=c(
  expression(mu ~-3*sigma), expression(mu ~-2 *sigma), expression(mu~-1 *sigma),  
  expression(mu), 
  expression(mu ~+1*sigma), expression(mu ~+2 *sigma), expression(mu ~+3 *sigma)))

#Compute probabilities
Fx <- pnorm(-4:4, mu, sigma)

#For 4 standard deviations either side of the mean
for(i in -4:4){
  j <- i+1
  k <- 5
  ijvals <- seq(i, j, 0.001)
  
  #Shade areas
  polygon(x = c(i,ijvals , j), 
          y = c(0, dnorm(ijvals, 0, 1), 0), 
          col=cols[i+k], border="white")
  
  #Add probability of each area as text
  text((i+0.5), 0.05, labels=round(diff(Fx)[i+k],digits=3), col=colText[i+k])
}
```

### Properties of the Normal Distribution

The normal distribution is a commonly used distribution. It is referred to as being **bell**-shaped. It is symmetric about the mean $\mu$, with half the probability falling on either side of $\mu$. The form of the probability density function means that unlike some other distributions, the probabilities cannot be expressed with a closed-form function. Numerical integration techniques have been used to evaluate the cumulative distribution function for the normal distribution. The resulting probabilities for the standard normal distribution are provided in tables.

::: {.callout-note appearance="simple"}

## Properties

The **cumulative distribution function** is not available in closed form, but can be evaluated:

- Converting $X\sim N(\mu, \sigma^2)$ to a standard normal random variable $Z\sim N(0,1)$ and using tables.
- Using software such as `R`. In `R`, the distribution function can be calculated using the command: `pnorm(x, mu, sigma)`

\textbf{Properties}

- Area under the pdf $f(x)$ is equal to 1
- Symmetric:  area to left of mean $\mu$ = 0.5 = area to right of mean $\mu$ 
- To compute $P(X \leq x)$ find the area to the left of $x$.
- Because $X$ is a continuous random variable, $P(X=x) =0$.

:::

### Example 14: Normal Distribution {.unnumbered}

A machining operation produces steel shafts where diameters have a normal distribution with a mean of 1.005 inches and a standard deviation of 0.01 inch. (Adapted from Scheaffer and Young, 2010, ex 5.87, page 351)

a) What is the probability that a randomly selected steel shaft will be less than 1.005 inches in diameter?

*Solution*

Let $X$ denote the diameter of steel shafts, where $X \sim N(1.005, 0.01)$.

$$P(X \leq 1.005) = P(X \leq \mu) = 0.5$$

```{r}
pnorm(1.005, 1.005, 0.01)
```

b) What is the probability that a randomly selected steel shaft will be less than 1.02 inches in diameter?

(i) estimate the probability using the properties of the normal distribution, then (ii) use the relevant R output below to determine the probability.

*Solution*

$$P(X \leq 1.02) \approx P(X \leq \mu + 1.5\sigma) \approx 0.5 + 0.341 + 0.7 (0.136)$$

```{r}
0.5 + 0.341 + 0.7*(0.136)
pnorm(1.02, 1.005, 0.01)
```

```{r echo=FALSE, results="hide"}
options(scipen = 8, digits = 5)
options(formatR.arrow=TRUE,width=60)
```

```{r}
pnorm(seq(0.95, 1.05, 0.01), 1.005, 0.01)
dnorm(seq(0.95, 1.05, 0.01), 1.005, 0.01)
pnorm(seq(0.95, 1.05, 0.01), 0.01, 1.005)
```

c) Specifications call for diameters to fall within the interval $1.00 \pm 0.02$ inches.
What percentage of the output of this operation will fail to meet specifications?

*Solution*

$$P(X \leq 0.98, X > 1.02) = 0.073 = 7.3\%$$

```{r}
pnorm(0.98, 1.005, 0.01)
1-pnorm(1.02, 1.005, 0.01)

pnorm(0.98, 1.005, 0.01)+1-pnorm(1.02, 1.005, 0.01)
```

d) What should be the mean diameter of the shafts produced to minimize the fraction that fail to meet specifications?

*Solution*

**1 inch**

e) Numerically verify your answers to a) -- c). 

*Solution*

There are several ways to estimate the probabilities. These are demonstrated below.

```{r}
x <- rnorm(1e5, 1.005, 0.01)

# a)
prop.table(table(x < 1.005))

# b)
sum(x < 1.02)/1e5

# c) 
prop.table(table(x > 1.02 | x < 0.98))

```